---
title: "eBird Occupancy Model Testing"
author: "Laura Graham"
date: "July 21, 2016"
output: 
  pdf_document:
    number_sections: true
---
```{r setup, message=FALSE, include=FALSE}
library(R2jags)
library(snowfall)
library(abind)
library(R2WinBUGS)
library(dplyr)
library(tidyr)
library(ggplot2)
load("locations_sml.rda")
load("locations_full.rda")
```

# Data preparation
Currently working with the data for hummingbirds from Colorado. The dataset has been generated in Python and I have filtered it such that only locations which have greater than 3 observations in at least one month are included. This reduces the number of locations from `r nrow(locations_full)` to `r nrow(locations_sml)`. 

The final data output from Python is a 4-dimensional array where the dimensions are: Year, Location, Species and Replicate. 

```{r load_data}
load("data/hummingbirds_colorado.rda")

# size of array
dim(all_dat$sp_obs)
```

# Initial models
## Basic dynamic occupancy model (no covariates)

Firstly, we want to fit a model with no covariates and get some initial parameter estimates out.

```{r basic_model, eval = FALSE}
# Code modified from Kery and Schaub 2012
# Bundle data
all_dat$sp_obs[is.nan(all_dat$sp_obs)] <- NA
dat <- list(y = all_dat$sp_obs, nyear = dim(all_dat$sp_obs)[1], nsite = dim(all_dat$sp_obs)[2], nspecies = dim(all_dat$sp_obs)[3], nrep = dim(all_dat$sp_obs)[4])

# Initial values
zst <- apply(all_dat$sp_obs, c(1, 2, 3), max, na.rm=TRUE)	# Observed occurrence as inits for z
zst[is.infinite(zst)] <- NA
inits <- function(){ list(z = zst)}
# Parameters monitored
params <- c("psi", "phi", "gamma", "p")

# MCMC settings
ni <- 2500
nt <- 4
nb <- 500
nc <- 3

# Call JAGS from R
strt <- Sys.time()

source("code/JAGSParallel.R")
out <- JAGSParallel(n.cores = 3, data = dat, inits = inits, params = params, model.file = "code/dynocc.JAGS.R", n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb)

save(out, file="basic_model.rda")

print(Sys.time() - strt)
print(out$JAGSoutput, digits=3)
```

## Dynamic occupany model correcting for detection bias. 
This model has covariates on the observation model and uses indicator variable selection to identify important parameters. Full set of parameters included is:

- time
- effort_hrs
- day
- year
- population per sq mile
- housing density
- number of observers

```{r covs_model, eval = FALSE}
# Code modified from Kery and Schaub 2012
# Bundle data
all_dat$sp_obs[is.nan(all_dat$sp_obs)] <- NA # data comes out of SQL/Python with missing values coded as NaN - causes issues in JAGS
nyear <- dim(all_dat$sp_obs)[1]
nsite <- dim(all_dat$sp_obs)[2]
#nsite <- 40 # make it run quicker for initial testing
nspecies <- dim(all_dat$sp_obs)[3]
nrep <-  dim(all_dat$sp_obs)[4]
#nrep <- 10 # again, quicker for testing 

# create the scaled parameters
# need a function to scale an n-dimensional array
scale_array <- function(in_dat) {
  m <- mean(in_dat, na.rm=TRUE)
  s <- sd(in_dat, na.rm=TRUE)
  if(is.null(dim(in_dat))) {
    out_dat <- (in_dat-m)/s
  } else {
    out_dat <- apply(in_dat, 1:length(dim(in_dat)), function(x) (x-m)/s)
  }
  return(out_dat)
}

param_dat <- all_dat[-1]
param_dat$year <- 1:nyear
param_dat <- lapply(param_dat, scale_array)
nparam <- length(param_dat)
dat <- list(y = all_dat$sp_obs, nyear = nyear, nsite = nsite, nspecies = nspecies, nrep = nrep, effort_hrs = param_dat$effort_hrs, day = param_dat$day, time = param_dat$time, number_observers = param_dat$number_observers, pop00_sqmi = param_dat$pop00_sqmi, housing_density = param_dat$housing_density, year = param_dat$year, nparam = nparam)

# Initial values
zst <- apply(all_dat$sp_obs, c(1, 2, 3), max, na.rm=TRUE)	# Observed occurrence as inits for z
zst[is.infinite(zst)] <- NA
#zst <- zst[,1:40,]
inits <- function(){ list(z = zst)}
# Parameters monitored
params <- c("psi", "phi", "gamma", "p", "alphap", "betap", "wp")

# MCMC settings
ni <- 15000
nt <- 4
nb <- 10000
nc <- 3

# Call JAGS from R
strt <- Sys.time()

source("code/JAGSParallel.R")
out <- JAGSParallel(n.cores = 3, data = dat, inits = inits, params = params, model.file = "code/dynocc_covs.JAGS.R", n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb)

save(out, file="covs_model.rda")

print(Sys.time() - strt)
```

```{r covs_analyse}
load("covs_model.rda")

# create data frame of summary statistics for analysis
jags_sum <- as.data.frame(out$JAGSoutput$summary) %>%
  mutate(param = rownames(.))

# get estimates for the indicator variables
wp <- filter(jags_sum, grepl('wp', param)) %>%
  mutate(param = substr(param, 4, nchar(param) - 1)) %>%
  separate(param, c("parameter", "species"), sep="\\,") %>%
  mutate(var_use = ifelse(mean > 0.5, 1, NA)) %>% select(parameter, species, var_use)

# get the parameter estimates for the 'important' variables
beta <- filter(jags_sum, grepl('betap', param)) %>%
  mutate(param = substr(param, 7, nchar(param) - 1)) %>%
  separate(param, c("parameter", "species"), sep="\\,") %>%
  select(mean, parameter, species) %>%
  merge(wp) %>%
  mutate(mean = mean*var_use)

# plot coefficient estimates in a covariate x species heatmap
ggplot(data = beta, aes(x=parameter, y=species, fill=mean)) + 
  geom_tile() + scale_fill_gradient2(low = "blue", high = "red")

p <- filter(jags_sum, grepl('p\\[', param))
